---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Thompson, Mitchell"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


-----

**EXECUTIVE SUMMARY: **
***Abalones are an economic and recreational resource that is threatened by a variety of factors which include: pollution, disease, loss of habitat, predation, commercial harvesting, sport fishing and illegal harvesting. Environmental variation and the availability of nutrients affect the growth and maturation rate of abalones. Over the last 20+ years it is estimated the commercial catch of abalone worldwide has declined in the neighborhood of 40%. Abalones are easily over harvested because of slow growth rates and variable reproductive success. Being able to quickly determine the age composition of a regional abalone population would be an important capability. The information so derived could be used to manage harvesting requirements. An observational study of a sample size of 1036 abalones considered to be representative of a larger population was undertaken. The intent was to predict the age of abalone from physical measurements and thus avoiding the necessity of counting growth rings for aging. Ideally, a growth ring is produced each year of age. Current methods of age determination are accomplished by drilling the shell and counting the number of shell rings using a microscope. This is a difficult, time consuming, and expensive process. The study ultimately failed and a subsequent effort to determine why was launched. Findings of this data exploration discovered probable reasons of study failure which include normalization of gender identifications in later stages of life, weight and volume loss during later stages of life, near equivalent values of median weight and volume in mid- to later stages of life, and extreme outliers among the infant class within the sample.*** 


-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(dplyr)
library(reshape2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("/Users/mk1989f_a/NWU Google Drive/MSDS401/Working/abalones.csv", sep = ",")

# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

(1)(a) (2 points) Use *summary()* to obtain and present descriptive statistics from mydata.  Use table() to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}

#quick inspection of data
summary(mydata)

#generate table1 using features CLASS and RINGS
table.class_by_rings <- table(mydata$CLASS, mydata$RINGS, dnn = c("CLASS", "RINGS"))
table.class_by_rings

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: A preliminary review of sample data on Abalones show a data frame consisting of 1036 samples having eight variables: Sex, Length, Diameter, Height, Whole weight in grams, Shuck size, Ring count, and a Class factor of A1 thru A5, denoting age with A1 being the youngest. A five-number summary and frequency table of Class by Rings depict a distribution among the SEX attribute that appears to have a slight left skew and a right skewness to the CLASS attribute. Additionally, there is a potential for outliers in the attributes WHOLE, SHUCK, and VOLUME as the maximum values for these categories look substantially higher than the 3rd quartile. The amount of skewness and level of outliers may impact generation of hypothesis. ***



(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}

#generate table2 using features SEX and CLASS
table.sex_by_class <- table(mydata$SEX, mydata$CLASS, dnn = c("SEX", "CLASS"))
#add margins
addmargins(table.sex_by_class)

barplot(table.sex_by_class, main = "SEX by CLASS Membership", beside = TRUE, legend = c("Female", "Infant", "Male"), col = c("darkblue", "darkred", "darkgreen"))

```

**Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: According to the provided sample and resultant data, Abalones belonging to the Class A1 (youngest), largely correlate to the Sex = Infant abalones, hereafter referred to as “Sex: I”. As a result of difficulty in assigning gender until later in life, the count of Sex: I abalones increases until about mid-life, Class A3, after which the counts decrease. Of particular note from the sample data is that as an abalone ages, or progresses through the Class attribute, the gender assignment achieves a state of near equilibrium beginning at mid-age. This may serve to be a hinderance to age prediction from physical measurements if gender differences are prevalent and not accounted for. This is also an important ratio to bear in mind when ascertaining population survivability ratios. ***




(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}

#set seed and obtain row indices for random sample of 200 without replacement
set.seed(123)
work <- sample(1:nrow(mydata), size = 200, replace = FALSE)
#draw random sample, assign to work
work <- mydata[work, ]

#construct scatterplot of columns 2-6 from sample
plot(work[, 2:6])

```

-----

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}

plot1 <- ggplot(data = mydata, aes(x = VOLUME, y = WHOLE)) + 
  geom_point(aes(color = CLASS), size = 1) + 
  labs(title = "Abalones:", subtitle = expression("Whole weight, in grams | Volume, cm"^"3"),
       x = "Volume", y = "Whole Weight", caption = "")
  
plot1 + theme(
  plot.title = element_text(size = 18, face = "bold.italic", hjust = 0.5),
  plot.subtitle = element_text(size = 16, face = "bold", hjust = 0.5),
  plot.caption = element_text(size = NULL),
  axis.title.x = element_text(face = "bold", size = 12),
  axis.title.y = element_text(face = "bold", size = 12)
)

```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}

plot2 <- ggplot(data = mydata, aes(x = WHOLE, y = SHUCK)) + 
  geom_point(aes(color = CLASS), size = 2, alpha = 0.5) + 
  labs(title = "Abalones:", subtitle = expression("Shuck weight, in grams | Whole weight, in grams"),
       x = "Whole Weight", y = "Shuck Weight", caption = "") +
  geom_abline(linetype = 2, color = "red", intercept = 0, slope = max(mydata$SHUCK / mydata$WHOLE))

plot2 + theme(
  plot.title = element_text(size = 18, face = "bold.italic", hjust = 0.5),
  plot.subtitle = element_text(size = 16, face = "bold", hjust = 0.5),
  plot.caption = element_text(size = NULL),
  axis.title.x = element_text(face = "bold", size = 12),
  axis.title.y = element_text(face = "bold", size = 12),
  legend.position = c(.98, .05),
  legend.justification = c("right", "bottom")
)


```

**Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: A greater range in variability exists between Whole weight vs. Volume than Shuck weight vs. Whole weight. It should also be noted that, on average, the total Shuck weight makes up less than half of the total Whole weight. This relationship also degrades over time. A slope and intercept line represent the maximum ratio of Shuck weight to Whole weight in plot 2(b). Deployment of this line in relation to the total Whole weight shows that most of the sampled abalones fall below the maximum ration, which could indicate that older abalones, in Classes A3 - A5, may not yield more shuck weight. Such a characteristic could impact age estimation if a researcher assumes a 1:1 ratio and does not account for the reduction in Whole weight as life progresses.***

-----

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}

hist <- ggplot(data = mydata, aes(x = RATIO, fill = SEX)) +
  geom_histogram(bins = 14, color = "black") +
  facet_wrap(~ SEX, nrow = 1)

box <- ggplot(data = mydata, aes(x = SEX, y = RATIO)) +
  geom_boxplot(aes(fill = SEX)) +
  facet_wrap(~ SEX, nrow = 1)

qq <- ggplot(mydata, aes(sample = RATIO, color = SEX)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ SEX, nrow = 1)
#plot3.3 <- 


#facet_grid(cols = vars(whatever variable you are breaking something down by))
#or
grid.arrange(hist, box, qq, nrow = 3)

```

**Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: According to the Ratio variable, the sampled abalone data frame result in non-normal for each gender: Infant, Male, and Female. Each histogram of count vs. Ratio by Sex conveys a skewed likely due to outliers per sex. This is particularly noticeable in the Female histogram. Complimenting the histograms and boxplots, Quantile-Quantile plots for the Infant and Female gender assignments each depict a tendency to drift up and away from normal, more so than the Male, further indicating a positive skewness of the distribution.***



(3)(b) (2 points) Use the boxplots to identify RATIO outliers (mild and extreme both) for each sex. Present the abalones with these outlying RATIO values along with their associated variables in "mydata" (Hint:  display the observations by passing a data frame to the kable() function).  

```{r Part_3b}

outliers <- boxplot.stats(mydata$RATIO)$out
kable(mydata[mydata$RATIO %in% outliers, ], table.attr = "style = \"color: black;\"")

```

**Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: One extreme outlier in the Sex: I, Class: A1, indicates a high number of rings (6) as well as a very large volume. Has this particular abalone been misidentified? Additional observations identified as outliers invovle the Sex: I, Class A1 more so than other genders. This aspect could hinder age prediction. Furthermore, each outlier involving the Sex: M is comprised of the Class: A3, or mid-life stage. Is this due to an increased rate of early development in the Sex: M gender group? Lastly, larger values of Whole weight and Volume are more closely related to older abalones and seem to reduce with age as previously mentioned.***

-----

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}

par(mfrow = c(2,2))
box1 <- boxplot(VOLUME ~ CLASS, data = mydata, xlab = "Class", ylab = "Volume")
box2 <- boxplot(WHOLE ~ CLASS, data = mydata, xlab = "Class", ylab = "Whole")
scat1 <- plot(VOLUME ~ RINGS, data = mydata, xlab = "Rings", ylab = "Volume")
scat2 <- plot(WHOLE ~ RINGS, data = mydata, xlab = "Rings", ylab = "Whole")
```

**Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: Boxplots indicate that deltas in median values of Volume and Whole weight become smaller as the age of the sampled abalones increases, with peak medians occurring in the Class: A4. These minor changes in median values are even more pronounced when comparing Class: A3 with Class: A5. This aspect of similar median values, especially at mid-life and full maturity, rightfully precludes it from being a stand-alone predictor of age by physical measurements.***

-----

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}

df <- aggregate(mydata[,c('VOLUME', 'SHUCK', 'RATIO')], by = list(mydata$SEX, mydata$CLASS), FUN = mean, na.rm = T)

colnames(df) <- c('SEX', 'CLASS', 'VOLUME', 'SHUCK', 'RATIO')
df

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}

r_vs_c <- ggplot(df, aes(CLASS, RATIO, group = SEX, colour = SEX)) +
  geom_line(lwd = 1.5) +
  geom_point(size = 3.5) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = 'Ratio by Class and Sex')

v_vs_c <- ggplot(df, aes(CLASS, VOLUME, group = SEX, colour = SEX)) +
  geom_line(lwd = 1.5) +
  geom_point(size = 3.5) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = 'Volume by Class and Sex')


s_vs_c <- ggplot(df, aes(CLASS, SHUCK, group = SEX, colour = SEX)) +
  geom_line(lwd = 1.5) +
  geom_point(size = 3.5) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = 'Shuck by Class and Sex')

grid.arrange(r_vs_c, v_vs_c, s_vs_c, nrow = 1)

```

**Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: From the sample set, it appears that after Class: A4, there is no clear distinction of Volume or Shuck weight being reliable to predict the gender. This is further supported and possibly enhanced by the Ratio variable in that there may be little distinction after Class: A2. ***



5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}

i_volring <- mydata %>%
  filter(SEX == "I", RINGS < 16) %>%
  ggplot(aes(x = RINGS, y = VOLUME, group = RINGS)) +
  geom_boxplot() +
  ylim(0, 1100) +
  theme_classic()

adult_volring <- mydata %>%
  filter(SEX %in% c("M", "F"), RINGS < 16) %>%
  ggplot(aes(x = RINGS, y = VOLUME, group = RINGS)) +
  geom_boxplot() +
  ylim(0, 1100) +
  theme_classic()

i_wholering <- mydata %>%
  filter(SEX == "I", RINGS < 16) %>%
  ggplot(aes(x = RINGS, y = WHOLE, group = RINGS)) +
  geom_boxplot() +
  ylim(0, 400) +
  theme_classic()

adult_wholering <- mydata %>%
  filter(SEX %in% c("M", "F"), RINGS < 16) %>%
  ggplot(aes(x = RINGS, y = WHOLE, group = RINGS)) +
  geom_boxplot() +
  ylim(0, 400) +
  theme_classic()

grid.arrange(i_volring, adult_volring, i_wholering, adult_wholering, ncol = 2)

```

**Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: Side by side boxplots of Volume vs Rings and Weight vs. Rings suggest a higher degree of range in Volume after gender assignment, which generally occurs in age classes A2 and A3 (see chart: Sex by Class Membership). Despite limiting the number of rings to less than 16, the graphs show that there is little difference in Whole Weight or Volume between Adults or Infants.***

-----

**Conclusions**

Please respond to each of the following questions (10 points total):

**Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: Presence of and not accounting for outliers may provide some indication to the failure of the original study. It should also be noted that, on average, the total Shuck weight makes up less than half of the total Whole weight. This relationship also degrades over time. Additionally, minor changes in median values are pronounced when comparing Class: A3 with Class: A5. This aspect of similar median values, especially at mid-life and full maturity, rightfully precludes it from being a stand-alone predictor of age by physical measurements. Physical measurements as an indicator of Age can be utilized as a launch point for further analysis and testing. Predictions and or conclusions should not be solely based on observational measurements, as such measurements are very prone to bias. Perhaps additional samples could further refine what limits should be placed on abalone age prediction based on physical measurements.***

**Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: Questions to ask would include, but not be limited to: How mature is the study? What is the sampling methodology? Is the methodology an accepted practice and/or peer reviewed? Where was the sample obtained? When was the sample obtained? Why and by whom was the sample attained? What is the sample size relative to the population? Does the study suffer from too little sampling? Over sampling? Who is funding the sampling and research respectively? Is there a conflict of interest? Does the study suffer from too little funding? Too much funding? Finally, what is the hypothesis before, after, and can it be statistically supported? ***

**Question 3)  (2 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: Observational studies possess an inherit risk in controlling measurement errors and sampling or researcher biases. Outliers also tend to increase the level of difficulty in the development of irrefutable conclusions. Additionally, observational studies are usually very limited in scope and do not consider factors that influence the articles being studied such as environmental, physical, physiological, or spatial and atmospheric contexts. Many studies have shown causality based on coincidence or simultaneous events from independent topical areas. Observational studies are generally not random either with regard to other factors that cannot be controlled. Despite of their risks, observational studies can and should serve as a form of hierarchy from simpler to more complex experiment designs and hypothesis generation. ***
