---
title: "Data Analysis #2 Version 2 (75 points total)"
author: "Thompson, Mitchell"

output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

-----

**EXECUTIVE SUMMARY: **
***Abalones are an economic and recreational resource that is threatened by a variety of factors which include: pollution, disease, loss of habitat, predation, commercial harvesting, sport fishing and illegal harvesting. Environmental variation and the availability of nutrients affect the growth and maturation rate of abalones. An observational study of a sample size of 1036 abalones considered to be representative of a larger population was undertaken with an intent to predict the age from physical measurements and thus avoiding the necessity of counting growth rings for aging. Data science methodologies were executed against the study's data set in two parts. Part I incorporated a data exploration effort in which findings discovered probable reasons of study failure which include normalization of gender identifications in later stages of life, weight and volume loss during later stages of life, near equivalent values of median weight and volume in mid- to later stages of life, and extreme outliers among the infant class within the sample. This report encompasses Part II and is built upon the explorative discoveries in attempts to discover and calculate variables that have statistical significance on predictive value in abalone harvesting. As a result of this analysis, three harvest strategies were developed at a 95% confidence level: Maximum Difference, Zero A1 Infants, and Equal Error. To achieve the highest harvest yield per proportion of 67.6%, the harvest would include all classes except A1 infants. However, the Zero A1 Infants strategy also involves the highest rate of error and the lowest volume. Conversely, the Max Difference strategy enjoys having the lowest false positive rate but carries with it the lowest proportional yield. The third strategy, Equal Error, splits the difference with a 62.4% proportional yield as well as a false positive error that resides comfortably in the middle. *** 

-----

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(flux)
library(ggplot2)
library(gridExtra)
library(moments)
library(rockchalk) 
library(car)
library(tidyverse)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("/Users/mk1989f_a/NWU Google Drive/MSDS401/Working/abalones.csv", sep = ",")
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```


###Section 1 (5 points)
(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a}

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO  <- mydata$SHUCK / mydata$VOLUME
s  <- rockchalk::skewness(mydata$RATIO)
k  <- rockchalk::kurtosis(mydata$RATIO, excess = FALSE)
k2 <- rockchalk::kurtosis(mydata$RATIO)

par(mfrow = c(1, 2))

hist(mydata$RATIO, main = "Histogram of Shuck / Vol Ratio", col = "cadetblue", xlab = "Ratio")
qqnorm(mydata$RATIO, main = "Q-Q Plot", col = "burlywood4")
qqline(mydata$RATIO)

```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b}

mydata$L_RATIO <- log10(mydata$RATIO)
s  <- rockchalk::skewness(mydata$L_RATIO)
k  <- rockchalk::kurtosis(mydata$L_RATIO, excess = FALSE)
k2 <- rockchalk::kurtosis(mydata$L_RATIO)

par(mfrow = c(1, 2))

hist(mydata$L_RATIO, main = "Histogram of Log Ratio", col = "deepskyblue", xlab = "Log Ratio")
qqnorm(mydata$L_RATIO, main = "Q-Q Plot", col = "burlywood4")
qqline(mydata$L_RATIO)

par(mfrow = c(1, 1))

boxplot(mydata$L_RATIO ~ mydata$CLASS, main = "Boxplots of Log Ratio by Class", ylab = "Frequency",
        xlab = "Distribution by Class", beside = TRUE, col = 'limegreen', outcol = "red")

```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}

bartlett.test(RATIO ~ CLASS, data = mydata)
bartlett.test(L_RATIO ~ CLASS, data = mydata)

```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: The L_RATIO variable exhibits better conformance to a normal distribution with a homogeneity of variance that remains nearly intact across the age classes. This is observed in the reduced skewness in the QQ Plot and histogram as well as even distribution of outliers in the QQ plot and boxplots. Additionally, bearing in mind that the Bartlett test is sensitive to deviations from normality, the test results depict the shape of a data set that leads to uphold the null hypothesis.***


###Section 2 (10 points)
(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}

summary(aov(L_RATIO ~ CLASS + SEX + CLASS:SEX, data = mydata))
summary(aov(L_RATIO ~ CLASS + SEX, data = mydata))

```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: The non-significant interaction term, 'CLASS:SEX', elevated the individual p-value score of the factor SEX by 0.00004 with a non-noticeable change to the individual factor CLASS. Additionally, the p-value of the interaction term itself, 0.86709, is greater than the scientific standard of significance and therefore is not a significant factor on the relationship. ***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}

tuk <- TukeyHSD(aov(L_RATIO ~ CLASS + SEX, data = mydata))
tuk

plot(tuk)

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: Using the scientific accepted standard p-value of less than 0.05, coefficients trending in the L_RATIO variable indicates significant differences in nearly every class comparison with the exception of A2-A1 only. This strongly suggests the L_RATIO variable declines with age. Such a finding supports earlier data exploration results pertaining to the relationship between RATIO and ageing associated with all three gender categories. Additionally, acquiring a high p-value and a low differential score in the male - female comparison suggests that a difference between the two is not statically significant. Therefore, the two gender categories can be combined into a single adult group for further analysis.***


###Section 3 (10 points)
(3)(a1) We combine "M" and "F" into a new level, "ADULT". (While this could be accomplished using *combineLevels()* from the 'rockchalk' package, we use base R code because many students do not have access to the rockchalk package.) This necessitated defining a new variable, TYPE, in mydata which had two levels:  "I" and "ADULT". 

```{r Part_3a1}
# here we show how to define the new variable TYPE using only base R functions (no need for outside packages)
mydata$TYPE <- character(nrow(mydata))  # initialize the TYPE column as all blanks
for (i in seq(along = mydata$SEX)) {
  mydata$TYPE[i] <- 'I'
  if (mydata$SEX[i] == 'M' || mydata$SEX[i] == 'F') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on definition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
table(mydata$SEX, mydata$TYPE)

```
(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2}

mydata$TYPE <- rockchalk::combineLevels(mydata$SEX, levs = c("F", "M"), "ADULT")

par(mfrow = c(1, 2))

hist(mydata$VOLUME[mydata$TYPE == 'I'], main = "Infant Volumes", col = "deepskyblue", xlab = "Volume", ylim = c(0, 120))
axis(side = 1, at = seq(0, 1000, by = 50), labels = FALSE, tcl = -0.2)
axis(side = 2, at = seq(0, 120, by = 5), labels = FALSE, tcl = -0.2)
hist(mydata$VOLUME[mydata$TYPE == 'ADULT'], main = "Adult Volumes", col = "forestgreen", xlab = "Volume", ylim = c(0, 150))
axis(side = 1, at = seq(0, 1000, by = 50), labels = FALSE, tcl = -0.2)
axis(side = 2, at = seq(0, 150, by = 10), labels = FALSE, tcl = -0.2)

```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: A preponderance of the distributed counts among the infants is within the lower volumes, leading to a right skewed distribution with a few outliers in the very high volumes. Meanwhile, the distribution among the combined adult group, comprised of both male and female variables, is close to normal. More specifically, distribution among the adult volume peaks between 250-450 compared to the peak in infant volume at approximately 175. While this does not necessarily equate to a difficulty, accounting for the volume will be a required part of the equation for separating the two classes of sex due to some overlap. ***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b}

mydata <- mydata %>% mutate(L_SHUCK = log10(SHUCK), L_VOLUME = log10(VOLUME))

shuck_vol <- ggplot(mydata, aes(VOLUME, SHUCK))
lshuck_lvol <- ggplot(mydata, aes(L_VOLUME, L_SHUCK))

grid.arrange(
  shuck_vol + geom_point(aes(fill = CLASS), shape = 21, alpha = 0.4),
  lshuck_lvol + geom_point(aes(fill = CLASS), shape = 21, alpha = 0.4),
  shuck_vol + geom_point(aes(fill = TYPE), shape = 21, alpha = 0.4),
  lshuck_lvol + geom_point(aes(fill = TYPE), shape = 21, alpha = 0.4)
)

```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: The log-transformation appears to increase the correlation coefficient when visually compared to the normal plot, thereby providing a stabilizing force to the variability. However, the model goodness of fit does not change as the distance away from the best fit line of all values is reduced by a factor of the log. Implications set forth by the plotted variables should provide a reliable predictor with increased chances of success during the linear regression testing. Finally, due to significant overlap in each normal plot, difficulty in determining levels by CLASS or TYPE is a challenge. This can be mitigated by the log-transformation plots which, for example, depicts the infant type having a larger grouping in the lower volumetric regions {<1.75, <0.75}.***


###Section 4 (5 points)
(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. Reclassify the infants in classes A4 and A5 as ADULTS. This reclassification could have been achieved using *combineLevels()*, but only on the abalones in classes A4 and A5. We will do this recoding of the TYPE variable using base R functions. We will use this recoded TYPE variable, in which the infants in A4 and A5 are reclassified as ADULTS, for the remainder of this data analysis assignment. 

```{r Part_4a1}
for (i in seq(along = mydata$TYPE)) {
  if (mydata$CLASS[i] == 'A4' || mydata$CLASS[i] == 'A5') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on redefinition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
cat('\nThree-way contingency table for SEX, CLASS, and TYPE:\n')
print(table(mydata$SEX, mydata$CLASS, mydata$TYPE))
```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}

model <- lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data = mydata)
summary(model)

```

**Essay Question:  Interpret the trend in CLASS level coefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: The estimated coefficients suggest a decrease in L_SHUCK for the higher CLASS variables, keeping L_VOLUME and TYPE constant. This is the expected trend as when compared to ealier charts, such as the L_VOLUME vs. L_SHUCK by CLASS, an inverse relationship appears as the variable L_SHUCK increases as L_VOLUME increases when plotted by CLASS. Thus, the pattern seems to a sharp increase in shuck values as one progresses through the classes, then flattens or begins to slightly decrease in the higher A4-A5 classes.***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: In this linear regression model, having a p-value of 0.006180 TYPE plays a significant role as a predictor. In a comparison of coefficient estimates among others used in this model TYPE does not seem to be as significant as any of the CLASS variables. ***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

### Section 5 (5 points)
(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5a}

s  <- rockchalk::skewness(model$residuals)
k  <- rockchalk::kurtosis(model$residuals, excess = FALSE)
k2 <- rockchalk::kurtosis(model$residuals)

par(mfrow = c(1, 2))
hist(model$residuals, main = "Histogram of Residuals", xlab = "Residuals", col = "deepskyblue")

qqnorm(model$residuals, main = "Q-Q Plot", col = "burlywood4")
qqline(model$residuals)

```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b}

mydata$RESIDUALS <- model$residuals
grid.arrange(
  ggplot(mydata, aes(L_VOLUME, RESIDUALS)) + 
    geom_point(aes(fill = CLASS), shape = 21, alpha = 0.5) +
    geom_smooth(aes(color = L_VOLUME, fill = L_VOLUME), size = .5,
                level = 0.95, method = "gam", formula = y ~ s(x, bs = "cs"), se = TRUE),
  ggplot(mydata, aes(L_VOLUME, RESIDUALS)) + 
    geom_point(aes(fill = TYPE), shape = 21, alpha = 0.5) +
    geom_smooth(aes(color = L_VOLUME, fill = L_VOLUME), size = .5,
                level = 0.95, method = "gam", formula = y ~ s(x, bs = "cs"), se = TRUE),
  ggplot(mydata, aes(CLASS, RESIDUALS)) +
    geom_boxplot(aes(fill = CLASS)),
  ggplot(mydata, aes(TYPE, RESIDUALS)) +
    geom_boxplot(aes(fill = TYPE))
)

bartlett.test(RESIDUALS ~ CLASS, data = mydata)

```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: From the calculations and subsequent graphical displays, the residuals appear to behave randomly which generates a strong suggestion that the model fits the data well. This is better illustrated by adding a generalized smoothing line, which as depicted on the scatter plots hovers at or near the 0.0 line indicating a good fit. Such a fit is quite revealing of how the variables L_VOLUME, and ultimately VOLUME will be useful for harvesting predictions and consequent decisions.  ***

-----

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

###Section 6 (5 points)
(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

# prop.infants shows the impact of increasing the volume cutoff for
# harvesting. The following code shows how to "split" the population at
# a 50% harvest of infants.

n.infants <- sum(prop.infants <= 0.5)
split.infants <- min.v + (n.infants + 0.5)*delta  # This estimates the desired volume.
split.infants

n.adults <- sum(prop.adults <= 0.5)
split.adults <- min.v + (n.adults + 0.5)*delta
split.adults

```

(6)(b) Present a plot showing the infant proportions and the adult proportions versus volume.value. Compute the 50% "split" volume.value for each and show on the plot.   

```{r Part_6b}

ggplot() +
  geom_line(aes(volume.value, prop.infants), color = "orange2") +
  geom_line(aes(volume.value, prop.adults), color = "deepskyblue") + 
  geom_vline(xintercept = split.infants, linetype = "dashed", color = "darkred") +
  annotate("text",
           label = paste(round(split.infants, 2), "\nInfants"),
           x = split.infants - 50, 
           y = 0.60,
           color = "orange2") +
  geom_vline(xintercept = split.adults, linetype = "dashed", color = "darkred") +
  annotate("text",
           label = paste(round(split.adults, 2), "\nAdults"),
           x = split.adults + 50, 
           y = 0.40,
           color = "deepskyblue") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "darkred") +
  labs(x = "Volume", y = "Proportion", title = "Protected Proportions of Adult and Infant Populations")

```

**Essay Question:  The two 50% "split" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: The 50-50 split values of 133.82 for infants and 384.51 for adults indicate a difference of 250.69 between the populations. This value suggests ample room among the volume between the populations that should allow for harvest at various cutoffs while protecting the repsective populations for sustainment.***

-----

This part will address the determination of a volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

###Section 7 (10 points)
(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}

mydiff <- ((1 - prop.adults) - (1 - prop.infants))
head(mydiff)

ggplot() +
  geom_line(aes(volume.value, mydiff)) +
  geom_vline(xintercept = split.infants, linetype = "dashed", color = "red") +
  annotate("text", label = paste('Infant Volume =', round(split.infants, 2)), 
           x = split.infants + 20, y = 0.2, angle = 90) +
  geom_vline(xintercept = split.adults, linetype = "dashed", color = "red") +
  annotate("text", label = paste('Adult Volume =', round(split.adults, 2)), 
           x = split.adults + 20, y = 0.2, angle = 90) +
  labs(title = "Difference in Adult and Infant Harvest Proportions",
       x = "Volume Value",
       y = "Difference in Proportions")

```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

```{r Part_7c}

mydiff1 <- volume.value[which.max(smooth.difference)]
ggplot() +
  geom_line(aes(volume.value, mydiff), color = "firebrick") +
  geom_line(aes(volume.value, smooth.difference), color = "cyan4", linetype = "dashed") +
  geom_vline(xintercept = mydiff1, linetype = "dashed", color = "black") +
  annotate("text", label = paste('volume =', round(mydiff1, 3)), x = mydiff1 + 20, y = 0.35, angle = 90) +
  labs(x = "VOLUME", y = "Difference in Proportions Harvested", 
       title = "Difference in Harvest Proportions")

```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions (NOTE:  the adult harvest proportion is the "true positive rate" and the infant harvest proportion is the "false positive rate").

Code for calculating the adult harvest proportion is provided.

```{r Part_7d}
#(1 - prop.adults)[which.max(smooth.difference)]  # [1] 0.7416332

tpr <- (1 - prop.adults)[which.max(smooth.difference)]
fpr <- (1 - prop.infants)[which.max(smooth.difference)]

paste("True positive rate:", tpr)
paste("False positive rate:", fpr)

```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

###Section 8 (10 points)
(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions.

```{r Part_8a}
base_cutoff <- volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786

a1.infant <- with(mydata, max(VOLUME[CLASS == 'A1' & TYPE == 'I']))
mydiff2 <- first(volume.value[volume.value > a1.infant])

paste("Base cutoff value:", base_cutoff)
paste("True positive rate:", (1 - prop.adults)[which(volume.value == mydiff2)])
paste("False positive rate:", (1 - prop.infants)[which(volume.value == mydiff2)])

```

(8)(b) Another cutoff is one for which the proportion of adults not harvested equals the proportion of infants harvested. This cutoff would equate these rates; effectively, our two errors:  'missed' adults and wrongly-harvested infants. This leaves for discussion which is the greater loss:  a larger proportion of adults not harvested or infants harvested?  This cutoff is 237.7383. Calculate the separate harvest proportions for infants and adults using this cutoff. Show these proportions.  Code for determining this cutoff is provided.  

```{r Part_8b}
proportional_cutoff <- volume.value[which.min(abs(prop.adults - (1-prop.infants)))] # [1] 237.6391

mydiff3 <- volume.value[which.min(abs(prop.adults + prop.infants - 1))]

paste("Proportional cutoff:", proportional_cutoff)
paste("True positive rate:", (1 - prop.adults)[which(volume.value == mydiff3)])
paste("False positive rate:", (1 - prop.infants)[which(volume.value == mydiff3)])

```


###Section 9 (5 points)
(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (7) and (8) on this plot and label each. 

```{r Part_9}

all_diffs <- which(volume.value %in% c(mydiff1, mydiff2, mydiff3))
ggplot(mapping = aes((1 - prop.infants), (1 - prop.adults))) + 
  geom_line(color = 'deepskyblue', size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  geom_point(mapping = aes((1 - prop.infants[all_diffs]), (1 - prop.adults[all_diffs])),
             shape = 3,
             size = 3,
             stroke = 1,
             color = c("red", "darkgreen", "black")) +
  annotate("text",
           label = paste('zero A1 infants. \nvol =', round(mydiff2, 1)), 
           x = 0.37,
           y = 0.76,
           size = 3,
           color = "red") +
  annotate("text",
           label = paste('equal harvest \nvol =', round(mydiff3, 1)), 
           x = 0.15,
           y = 0.85,
           size = 3,
           color = "darkgreen") +
  annotate("text",
           label = paste('max. difference \nvol =', round(mydiff1, 1)), 
           x = 0.25,
           y = 0.65,
           size = 3,
           color = "black") +
  labs(title = "ROC Harvest Proportions", x = "Infant Proportion", y = "Adult Proportion")
  
```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}

f <- flux::auc(x = (1 - prop.infants), y = (1 - prop.adults))
paste("Area under ROC curve:", f)

```


###Section 10 (10 points)
(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population
 	
```{r Part_10} 	

TPR <- (1 - prop.adults)[all_diffs]
FPR <- (1 - prop.infants)[all_diffs]
yield <- (TPR * total.adults + FPR * total.infants) / (total.adults + total.infants)

tribble(~strategy, ~volume, ~TPR, ~FPR, ~PropYield,
  "max.difference", mydiff1, TPR[3], FPR[3], yield[3],
  "zero.A1.infants", mydiff2, TPR[1], FPR[1], yield[1],
  "equal.error", mydiff3, TPR[2], FPR[2], yield[2]
)

```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the three cutoffs determined in this assignment.**   

***Answer: To achieve the highest harvest yield per proportion of 67.6%, the harvest would include all classes except A1 infants. However, the Zero A1 Infants strategy also involves the highest rate of error and the lowest volume. Conversely, the Max Difference strategy enjoys having the lowest false positive rate but carries with it the lowest proportional yield. The third strategy, Equal Error, splits the difference with a 62.4% proportional yield as well as a false positive error that resides comfortably in the middle. ***


**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for implementation of a cutoff?  4)  What suggestions would you have for planning future abalone studies of this type? 

***Answer: A presentation of this data analysis would first begin with any assumptions and limitations that were involved, followed by qualifications. To touch on these areas briefly, acknowledgement of weight and volume loss during later stages of life as well as near equivalent values of median weight and volume in mid- to later stages of life may present themselves as confusers in the analytical methods and should be accounted for. Additionally, the presence of extreme outliers among the infant class within the sample may suggest misidentification during the classification stages of data capture. The boxplot charts would be used during this portion of discussion to aid in visual understanding. Finally, a clear understanding by all parties that results of this analysis pertains to this data and its origins only. Another analysis of similar nature against another survey of abalones will end with different results. All of this portion of the discussion would be guided by the visual charts and graphs while reducing the amount of intimate analytical tradecraft relied upon. Immediately following the laydown of assumptions and limitations, multiple courses of action would be presented. Specific recommendations would not be made. Each course of action would entail the benefits of the selected strategy in terms of predicted harvest yield and health to the overall population. Also included would be the tradespace contained within the strategy relative to other strategies. Finally, if the investigative decision is to proceed with a harvest, a petition towards a more conservative approach would be levied in order to mitigate overharvesting and retain a sustainable growth. Suggestions for future studies of this type as well as future population characterization would be made at this time. Pointing back to the origination of outliers in this data set would be used as a discussion enabler. Another recommendation would be to consider external influences on abalone growth such as environmental, geographical, and dietary factors.  ***